---
title: "Get Herbarium Data"
format:
  html:
    toc: true
execute:
  cache: true
  freeze: auto
---

## What this script does

Download herbarium specimen records from four biodiversity databases:

- speciesLink  
- REFLORA  
- JABOT  
- GBIF  

The records are merged into a single dataset and exported as an **Excel file**.


## Setup

Load the R packages required to access biodiversity databases.

```{r eval = FALSE}

#install.packages("devtools")

# refloraR: https://dboslab.github.io/refloraR-website/
#devtools::install_github("DBOSlab/refloraR")
library(refloraR)

# jabotR: https://dboslab.github.io/jabotR-website/
#devtools::install_github("DBOSlab/jabotR")
library(jabotR)

# rspeciesLink: https://github.com/LimaRAF/plantR
# rgbif2: https://github.com/LimaRAF/plantR
#devtools::install_github("LimaRAF/plantR")
library(plantR)
```

Load general packages to operate data frames and export to **Excel**.

```{r eval = FALSE}

#install.packages(dplyr)
library(dplyr)

#install.packages(openxlsx)
library(openxlsx)
```

Load auxiliary functions for the script.

```{r eval = FALSE}

source("functions/auxiliary_functions.R")
```

Create folders in the directory to save the occurrence data.

```{r eval = FALSE}

if (!dir.exists("input_data/")) dir.create("input_data/", recursive = TRUE)
if (!dir.exists("output_data/")) dir.create("output_data/", recursive = TRUE)

timestamp_tag <- format(Sys.time(), "%d%b%Y")
all_data_path   <- file.path("output_data/", paste0("all_data_",  timestamp_tag, ".xlsx"))
```

Define the species names used to query the databases.

```{r eval = FALSE}
#Example:

targeted_taxa <- c(
  "Ormosia coarctata",
  "Ormosia stipularis",
  "Ormosia fastigiata"
)
```


## speciesLink

Download records from *speciesLink*.
This function requires a personal API key to download the data. You can generate one at:

*https://specieslink.net/aut/profile/apikeys*

```{r eval = FALSE}
#Example:

personal_key <- "DBpBRprOuNV8ZWoxRFcr"
```

To query multiple species, the function downloads a separate dataset for each species in the list.

```{r eval = FALSE}
data_speciesLink <- lapply(
  targeted_taxa,
  function(x) plantR::rspeciesLink(
    dir = "input_data/",
    filename = "spLink_records_search",
    save = TRUE,
    key = personal_key,
    basisOfRecord = NULL,
    family = NULL,
    species = x,
    collectionCode = NULL,
    country = NULL,
    stateProvince = NULL,
    county = NULL,
    Coordinates = NULL,
    CoordinatesQuality = NULL,
    Scope = NULL,
    Synonyms = "no synomyms",
    Typus = FALSE,
    Images = FALSE,
    RedList = NULL,
    MaxRecords = 5000,
    file.format = "csv",
    compress = FALSE
  )
)
```

Next, the resulting data frames are combined into a single dataset.

```{r eval = FALSE}

# Adding species names to each element of the list
names(data_speciesLink) <- targeted_taxa

# Binding all searchs and keeping a column with the named used in the search
data_speciesLink <- dplyr::bind_rows(data_speciesLink, .id = "original_search")
```

Finally, an auxiliary function standardizes the column names to conform to Darwin Core terminology.

```{r eval = FALSE}

data_speciesLink <- fix_column_names(data_speciesLink)
```

## REFLORA

Download specimen records from the *REFLORA Virtual Herbarium*.

```{r eval = FALSE}
data_reflora <- refloraR::reflora_records(
  herbarium = NULL,
  repatriated = TRUE,
  taxon = targeted_taxa,
  state = NULL,
  recordYear = NULL,
  indets = TRUE,
  reorder = c("herbarium", "taxa", "collector", "area", "year"),
  path = NULL,
  updates = TRUE,
  verbose = TRUE,
  save = FALSE,
  dir = "input_data/",
  filename = "reflora_records_search"
)
```

The *REFLORA* package temporarily stores individual downloaded records in a local directory. Since the records are already loaded into the `data_reflora` data frame, the temporary folder is removed to keep the workspace clean.

```{r eval = FALSE}

unlink("reflora_download/", recursive = TRUE, force = TRUE)
```

## JABOT

Download specimen records from the *JABOT* database of the *Rio de Janeiro Botanical Garden*.

```{r eval = FALSE}
data_jabot <- jabotR::jabot_records(
  herbarium = NULL,
  taxon = targeted_taxa,
  state = NULL,
  recordYear = NULL,
  indets = TRUE,
  reorder = c("herbarium", "taxa", "collector", "area", "year"),
  path = NULL,
  updates = TRUE,
  verbose = TRUE,
  save = FALSE,
  dir = "input_data/",
  filename = "jabot_records_search"
)
```

As in the *REFLORA* step, the temporary download directory is removed because the records are already stored in the `data_jabot` data frame.

```{r eval = FALSE}

unlink("jabot_download/", recursive = TRUE, force = TRUE)
```


## GBIF

Download occurrence records from the *Global Biodiversity Information Facility (GBIF)*.
As in the *speciesLink* query, the function downloads a separate dataset for each species in the list in a multi species search.

```{r eval = FALSE}
data_gbif <- lapply(
  targeted_taxa,
  function(x) plantR::rgbif2(
    dir = "input_data/",
    filename = "gbif_records_search",
    species = x,
    n.records = 5000,
    force = TRUE,
    remove_na = FALSE,
    save = FALSE,
    file.format = "csv",
    compress = FALSE
  )
)
```

The resulting data frames are then combined into a single dataset.

```{r eval = FALSE}

# Adding species names to each element of the list
names(data_gbif) <- targeted_taxa

# Binding all searchs and keeping a column with the named used in the search
data_gbif <- dplyr::bind_rows(data_gbif, .id = "original_search")
```


## Merge all sources

First, add a column identifying the origin of each record.

```{r eval = FALSE}
data_gbif$source <- "GBIF"
data_jabot$source <- "JABOT"
data_reflora$source <- "REFLORA"
data_speciesLink$source <- "speciesLink"
```

Then convert all fields to character before merging.

```{r eval = FALSE}

data_list <- list(data_gbif, data_jabot, data_reflora, data_speciesLink)

data_list <- lapply(data_list, function(df) {
  df %>% dplyr::mutate(across(everything(), as.character))
})
```

Combine the sources.

```{r eval = FALSE}

all_data <- dplyr::bind_rows(data_list)
```

# Export dataset

Export the merged dataset to Excel.

```{r eval = FALSE}

openxlsx::write.xlsx(all_data, file = all_data_path)
all_data_path
```



