---
title: "Clean Occurrence Data"
format:
  html:
    toc: true
execute:
  cache: true
  freeze: auto
---

## What this script does

Clean and standardize herbarium specimen records previously downloaded from biodiversity databases.The workflow uses the `barRoso` package, which standardizes biodiversity occurrence datasets and flags common issues such as duplicates, missing identifications, and problematic collections.

## Setup

Load the R packages required for taxonomic and dataset standardization.

```{r eval = FALSE}

#install.packages("devtools")

# LCVP: https://github.com/idiv-biodiversity/LCVP
#devtools::install_github("idiv-biodiversity/LCVP")
library(LCVP)

# lcvplants: https://github.com/idiv-biodiversity/lcvplants
#devtools::install_github("idiv-biodiversity/lcvplants")
library(lcvplants)

# barRoso: https://github.com/DBOSlab/barRoso
#devtools::install_github("DBOSlab/barRoso")
library(barRoso)
```

Load general packages to operate data frames and export to **Excel**.

```{r eval = FALSE}

#install.packages(dplyr)
library(dplyr)

#install.packages(openxlsx)
library(openxlsx)
```

Load auxiliary functions for the script.

```{r eval = FALSE}

source("functions/auxiliary_functions.R")
```

Load the merged dataset generated in the previous step of the pipeline, if it is not already present in the **R** environment.

```{r eval = FALSE}

all_data <- read.xlsx("output_data/")
```

Create folders in the directory to save the occurrence data:

```{r eval = FALSE}

if (!dir.exists("output_data/")) dir.create("output_data/", recursive = TRUE)

timestamp_tag <- format(Sys.time(), "%d%b%Y")
clean_data_path <- file.path("output_data/", paste0("clean_data_",  timestamp_tag, ".xlsx"))
```

## Harmonize dataset structure

Convert country names to standard English format using a auxiliary function.

```{r eval = FALSE}
if ("country" %in% names(all_data)) {
  all_data$country <- convert_country_EN(all_data$country)
}
```

Different biodiversity databases provide slightly different field structures.The following step harmonizes the dataset structure by selecting relevant fields and inserting missing columns.

```{r eval = FALSE}
cols <- c(names(data_reflora))

cols <- insert_after_word(cols, c(
  basisOfRecord = "gbifID",
  stateProvince = "county",
  scientificName = "acceptedScientificName",
  maximumElevationInMeters = "elevation",
  identificationRemarks = "vernacularName",
  identificationRemarks = "eventRemarks",
  fieldNotes = "habitat",
  day = "lifeStage",
  catalogNumber = "barcode"
))

all_data <- all_data %>%
  select(any_of(cols)) %>%
  filter(basisOfRecord %in% c("PRESERVED_SPECIMEN", "Preservedspecimen"))
```

## Clean the dataset

The `barroso_std()` function performs several cleaning and validation procedures for herbarium occurrence datasets. These procedures are controlled by function arguments and can be enabled depending on the user's objectives.

In this example, the function:

-   Flags possible misidentifications (`flag_missid`)

-   Detects duplicate specimens (`flag_duplicates`)

-   Removes duplicated records (`rm_duplicates`)

-   Identifies unvouchered records (`unvouchered`)

-   Removes problematic collections (`delunkcoll`)

Additional arguments and options can be explored in the function documentation using `?barroso_std()`.

```{r eval = FALSE}
clean_data <- barRoso::barroso_std(
  all_data,
  unvouchered = TRUE,
  delunkcoll = TRUE,
  flag_missid = TRUE,
  flag_duplicates = TRUE,
  rm_duplicates = TRUE,
  rm_original_column = TRUE
)
```

## Export the cleaned dataset

Export the cleaned dataset to **Excel** for manual inspection and correction.

```{r eval = FALSE}

openxlsx::write.xlsx(clean_data, file = clean_data_path)
}
```

## Reload the dataset after manual corrections (optional)

You may need to make manual corrections on the formatting of the collector names, for example. If needed, the `barroso_std()` function can be executed again after manual edits to ensure the dataset remains standardized.

```{r eval = FALSE}

clean_data <- read.xlsx("output_data/clean_data_20260112_160824.xlsx")
```
