---
title: "Add morphological traits"
format:
  html:
    toc: true
execute:
  cache: true
  freeze: auto
---

## What this script does

Generate a morphological trait matrix associated with the curated species occurrence dataset produced in previous steps of the pipeline.

The workflow uses the `morph_add_traits()` function of the `moRphoTaxa` package, which generates a standardized matrix of qualitative and quantitative morphological traits. The resulting dataset can include general traits shared across angiosperms as well as traits specific to selected plant groups — angiosperm families and, for non-angiosperm lineages (mosses, lycophytes, ferns and conifers), higher-rank groups.

The final dataset can be exported as an **Excel** file for data entry and morphometric analyses.

## Setup

Load the `moRphoTaxa` package, required to create and export data frames.

```{r eval = FALSE}

# moRphoTaxa: https://dboslab.github.io/moRphoTaxa-website/
#devtools::install_github("DBOSlab/moRphoTaxa")
library(moRphoTaxa)
```

If the curated occurrence dataset from the previous steps is not loaded in the **R** environment, it can be imported and used as the base data frame for the trait dataset. Otherwise, the workflow generates a trait matrix without associated species records.

```{r eval = FALSE}

clean_data <- read.xlsx("output_data/")
```

Create a folder to store the resulting dataset, and define the file path that will be passed to `morph_add_traits()` via `file_name`.

```{r eval = FALSE}

if (!dir.exists("output_data/")) dir.create("output_data/", recursive = TRUE)

timestamp_tag <- format(Sys.time(), "%d%b%Y")
trait_data_path <- file.path("output_data/", paste0("trait_data_",  timestamp_tag, ".xlsx"))
```

## Inspect available traits (optional)

`traits_database` is a reference data frame bundled with the package, listing every qualitative and quantitative trait currently implemented — including its full name, abbreviated code, trait type, and the trait set (`trait_scope`) it belongs to. Inspecting it before running `morph_add_traits()` helps decide which default and taxon-specific traits to request.

```{r eval = FALSE}

data(traits_database)

# available trait sets
unique(traits_database$trait_scope)

# traits for a given group, e.g. Passifloraceae
subset(traits_database, trait_scope == "passifloraceae")
```

## morph_add_traits()

The `morph_add_traits()` function creates a data frame containing morphological traits and optionally associates them with an occurrence dataset.

The traits included in the dataset are controlled through function arguments:

### **`base_df`**
A data frame containing species occurrence records. If `NULL`, the function creates only the trait matrix without associating it with occurrence data.

### **`trait_name`**
Defines how trait columns are labeled:

-   `"trait full name"` → descriptive trait names\
-   `"trait code"` → abbreviated trait codes\
-   `"both"` → full name and code combined

### **`quali_default` and `quanti_default`**
Logical arguments controlling whether the default qualitative and quantitative traits commonly used across angiosperms are included. Both are automatically switched to `FALSE` when `quali_specific`/`quanti_specific` refers to a non-angiosperm group (`"bryophyta"` or `"polipodiopsida"`), since the angiosperm default traits (corolla, calyx, fruit, seed, etc.) do not apply to those lineages. A message is printed when this happens, unless the argument was already set to `FALSE` or `verbose = FALSE`.

### **`quali_specific` and `quanti_specific`**
Character arguments used to add taxon-specific traits. Currently available presets include:

-   `"asteraceae"`
-   `"bryophyta"`
-   `"cactaceae"`
-   `"caesalpinioidae"`
-   `"orchidaceae"`
-   `"papilionoideae"`
-   `"passifloraceae"`
-   `"polipodiopsida"`

If `NULL`, only the default traits will be returned.

### **`save`**
Logical argument controlling whether the resulting dataset is exported as an **Excel** file.

### **`file_name`**
Character string with the file path (including file name and `.xlsx` extension) used to save the resulting data frame when `save = TRUE`. If `NULL`, a file is saved automatically inside `output_data/`, with a name generated from the current date and time. Ignored when `save = FALSE`.

### **`verbose`**
Logical argument controlling whether informative messages are printed during execution, including the location of exported files and any automatic `quali_default`/`quanti_default` adjustments for non-angiosperm presets.

## Add morphological traits to the dataset

In this example, the function:

-   Creates a morphological trait dataset\
-   Includes the default qualitative and quantitative traits\
-   Adds traits specific to *Papilionoideae* (Leguminosae)\
-   Exports the dataset as an Excel file at the path defined above

Users should modify the function arguments according to their needs (e.g., selecting different trait sets or plant groups).

```{r eval = FALSE}

df_morph_traits <- morph_add_traits(base_df = NULL,
                                    trait_name = "both",
                                    quali_default = TRUE,
                                    quanti_default = TRUE,
                                    quali_specific = "papilionoideae",
                                    quanti_specific = "papilionoideae",
                                    save = TRUE,
                                    file_name = trait_data_path,
                                    verbose = TRUE)
```

The output of the function is a data frame containing the original occurrence data (if provided) along with columns for the selected morphological traits; the dataset is also exported as an **Excel** file at `trait_data_path`, allowing manual insertion of trait values.

::: {.callout-note}
For a non-angiosperm dataset (e.g. mosses), requesting the corresponding preset directly disables the angiosperm defaults:

```{r eval = FALSE}

df_morph_traits_moss <- morph_add_traits(base_df = NULL,
                                         trait_name = "both",
                                         quali_specific = "bryophyta",
                                         quanti_specific = "bryophyta",
                                         save = TRUE,
                                         file_name = "output_data/trait_data_bryophyta.xlsx")
```

:::
